---
title: "Misk-DSI-GroupAssignment-1-SSH"
output: html_document
date: "2022-10-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig_width =10,echo = TRUE)
```

## Spotify Dataset Analysis

Spotify is a supplier of audio streaming and other media services that was established in October 2008 and is headquartered in Sweden. It is now one of the largest digital music, podcast, and video streaming services in the world, and it provides users with access to millions of songs created by artists from all over the world.

We start by uploading the important libraries and importing the data set under the investigation.
```{r load-packages, include=FALSE}

library(tidyverse)
library(ggplot2)
library(dplyr)
library(lubridate)
library(janitor)
library(tidytuesdayR)
library(ggcorrplot)
library(ggpubr)
```


```{r include=FALSE}
spotify_songs= read_csv(url("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-01-21/spotify_songs.csv"))
```

## Data Wrangling

We checked and identify the NA values, and then removed them to keep only the observations that have no missing values. 

```{r}
# Check NA values
which(is.na(spotify_songs), arr.ind=TRUE)

# Getting  the names of the R data frame columns that contain missing values.
spotify_songs1 = as.data.frame(cbind(lapply(lapply(spotify_songs, is.na), sum)))
rownames(subset(spotify_songs1, spotify_songs1$V1 != 0))

# Removing and keeping the clean data without NA values
Clean_data=na.omit(spotify_songs)
```

## Data Summary

In this section, we try to figure out some of general questions. 
For example what is the most top artist, album and gere?

```{r fig1, fig.align = "center", out.width = "70%"}
# Top 5 Artists
Top_Artists <- Clean_data %>%
               group_by(track_artist) %>%
               summarise(n_apperance = n()) %>%
               filter(n_apperance > 100) %>%
               arrange(desc(n_apperance))

Top_Artists$track_artist <- factor(Top_Artists$track_artist, levels = Top_Artists$track_artist [order(Top_Artists$n_apperance)])

ggplot(Top_Artists, aes(x = track_artist, y = n_apperance)) +
       geom_bar(stat = "identity", width = 0.6,fill = "grey", color="black") +
       labs(title = "Popular Artists", x = "Artists", y = "Number of Apperances in the list") +
       theme(plot.title = element_text(size=15, hjust=0, face = "bold"), axis.title = element_text(size=12)) +
       geom_text(aes(label=n_apperance), hjust=2, size = 3, color = 'white') +
       coord_flip()
```

We can see from the figure above that Martin Garrix is the most popular artist.

```{r fig2, fig.align = "center", out.width = "70%"}
# Top 5 Albums
Top_Albums <- Clean_data %>%
              group_by(track_album_name) %>%
              summarise(n_apperance = n()) %>%
              filter(n_apperance >= 29) %>%
              arrange(desc(n_apperance))

Top_Albums$track_album_name <- factor(Top_Albums$track_album_name, levels = Top_Albums$track_album_name [order(Top_Albums$n_apperance)])

ggplot(Top_Albums, aes(x = track_album_name, y = n_apperance)) +
      geom_bar(stat = "identity" , width = 0.6,fill = "grey", color="black") +
      labs(title = "Popular Albums", x = "Albums", y = "Number of Apperances in the list") +
      theme(plot.title = element_text(size=15, hjust=0, face = "bold"), axis.title = element_text(size=12)) +
      geom_text(aes(label=n_apperance), hjust=2, size = 3, color = 'white') +
      coord_flip()
```

We can observe that Greatest Hits fowlled by Ultimate Freestyle Mega Mix Gold are the most popular Albums.

```{r fig3, fig.align = "center", out.width = "70%"}
# The Most Popular Genre
ggplot(Clean_data) + 
         geom_bar(mapping = aes(x = playlist_genre), fill = "grey", color="black") + 
         labs(title = "The Most Popular Genre", x = "Playlist Genre", y = "Number of Genre")+
         theme(plot.title = element_text(size=15), axis.text = element_text(size = 15), axis.title = element_text(size = 15))
  
```

We can see clearly from the above figure that the most popular genre is edm.

So, we came with some general questions:

1. Is there any feature that increase the popularity of edm gene? 

To answer the question, we created the following boxplots 

```{r fig4, fig.align = "center", out.width = "70%"}
ggplot(Clean_data, aes(x=energy, y=playlist_genre)) + 
  geom_boxplot() +
  geom_boxplot(fill='grey', color="black")+
  theme_classic()
```

The plot shows that edm genre has songs with highest energy.

```{r fig5, fig.align = "center", out.width = "70%"}
ggplot(Clean_data, aes(x=danceability, y=playlist_genre)) + 
  geom_boxplot() +
  geom_boxplot(fill='grey', color="black")+
  theme_classic()
```

We also observe that rap genre has the highest danceability factor while latin genre has a higher valence than others.

```{r fig6, fig.align = "center", out.width = "70%"}
ggplot(Clean_data, aes(x=valence, y=playlist_genre)) + 
  geom_boxplot() +
  geom_boxplot(fill='grey', color="black")+
  theme_classic()

```

2. What is the most song features among listeners?

```{r fig7, fig.align = "center", out.width = "70%"}
energy_only <- cut(Clean_data$energy, breaks = 5)
Clean_data %>%
             ggplot( aes(x = energy_only )) +
             geom_bar( fill = "grey", colour = "black") +
             scale_x_discrete(name = "Energy")
```

This plot demonstrates that Spotify users favour songs with higher energy levels.

```{r fig8, fig.align = "center", out.width = "70%"}
speech_only <- cut(Clean_data$speechiness, breaks = 5)
Clean_data %>%
             ggplot( aes(x = speech_only )) +  
             geom_bar(fill = "grey", colour = "black") +
             scale_x_discrete(name = "Speechiness") 
```

In addtion, the above plot shows that maximum Spotify listeners prefer tracks with less vocals.

## Now let's test some hypothesis and see if the variables are indeed correlated.

```{r fig9, fig.height = 10, fig.width = 10, fig.align = "center", out.width = "70%"}
# Select only numerical variables
Clean_data_numeric= Clean_data %>% dplyr::select(where(is.numeric))
corr <- round(cor(Clean_data_numeric),3)
ggcorrplot(corr, hc.order = TRUE, type = "lower",
             lab = TRUE)
```

We found that there is a positive correlation between energy and loudness, while energy is negatively correlated with acousticness.

1. Is the Energy correlated with acousticness and loudness in specifc gere?

```{r fig10, fig.align = "center", out.width = "70%"}

# loudness & Energy.
val_nrgy <-Clean_data%>%ggplot(aes(x=loudness,y=energy))+geom_point()+ggtitle("Loudness/Energy")+theme_light()+geom_smooth(se=FALSE)+facet_wrap(~playlist_genre)
val_nrgy # More energetic song means the more loudness nearly in all genre

# acousticness & Energy.

val_nrgy1 <-Clean_data%>%ggplot(aes(x=acousticness,y=energy))+geom_point()+ggtitle("Acousticness/Energy")+theme_light()+geom_smooth(se=FALSE)+facet_wrap(~playlist_genre)
val_nrgy1 ## Energy & Acoustic. The lower the energy the more the acoustic.
```

We can see that there is a correlation between loudness and Energy in most gere and the same for acousticness and Energy.


2. Is there relationship between album release date and popularity?
3. Does some features affected by album release date?


```{r fig11, fig.align = "center", out.width = "70%"}
Y = Clean_data$acousticness
X = as.numeric(substr(Clean_data$track_album_release_date, 1,4))
by_years =data.frame(X,Y)
by_years <- aggregate(Y~X, data = by_years, mean)
g= ggplot(by_years, aes(x = X, y = Y)) +
  geom_line(size=1.4) +
  labs(title = "Acousticness - Years Graph",
       x = "Years",
       y = "Avg. Acousticness")
```

```{r fig12, fig.align = "center", out.width = "70%"}
Y1 = Clean_data$loudness
X1 = as.numeric(substr(Clean_data$track_album_release_date, 1,4))
by_years1 =data.frame(X1,Y1)
by_years1 <- aggregate(Y1~X1, data = by_years1, mean)
g1= ggplot(by_years1, aes(x = X1, y = Y1)) +
  geom_line(size=1.4) +
  labs(title = "Loudness - Years Graph",
       x = "Years",
       y = "Avg. Loudness")

```


```{r}
Y2 = Clean_data$danceability
X2 = as.numeric(substr(Clean_data$track_album_release_date, 1,4))
by_years2 =data.frame(X2,Y2)
by_years2 <- aggregate(Y2~X2, data = by_years2, mean)
g2 = ggplot(by_years2, aes(x = X2, y = Y2)) +
  geom_line(size=1.4) +
  labs(title = "Danceability - Years Graph",
       x = "Years",
       y = "Avg. Danceability")
```

```{r}
Y3 = Clean_data$speechiness
X3 = as.numeric(substr(Clean_data$track_album_release_date, 1,4))
by_years3 =data.frame(X3,Y3)
by_years3 <- aggregate(Y3~X3, data = by_years3, mean)
g3 = ggplot(by_years3, aes(x = X3, y = Y3)) +
  geom_line(size=1.4) +
  labs(title = "Speechiness - Years Graph",
       x = "Years",
       y = "Avg. Speechiness")
```



```{r}
Y4 = Clean_data$track_popularity
X4 = as.numeric(substr(Clean_data$track_album_release_date, 1,4))
by_years4 =data.frame(X4,Y4)
by_years4 <- aggregate(Y4~X4, data = by_years4, mean)
g4 = ggplot(by_years4, aes(x = X4, y = Y4)) +
  geom_line(size=1.4) +
  labs(title = "Popularity - Years Graph",
       x = "Years",
       y = "Avg. popularity")
```

```{r}
Y5 = Clean_data$energy
X5 = as.numeric(substr(Clean_data$track_album_release_date, 1,4))
by_years5 =data.frame(X5,Y5)
by_years5 <- aggregate(Y5~X5, data = by_years5, mean)
g5 = ggplot(by_years5, aes(x = X5, y = Y5)) +
  geom_line(size=1.4) +
  labs(title = "Energy - Years Graph",
       x = "Years",
       y = "Avg. Energy")

```


```{r fig13, fig.align = "center", out.width = "70%"}
figure <- ggarrange(g, g1, g2,g3, g4,g5, ncol = 3, nrow = 2)
figure
```

Unlike the acoustickness, we can notice that there is a major drop in 1960 for popularity, energy, danceability, loudness, and speechiness. However, we can see that new songs are less acoustic, speechines and popularity than the older ones while danceability and loudness have an increase trend with time. on the other hande energy became slightly stable after 1980.


## Conclusion 

We have performed data wrangling on our spotify dataset by removing null values before starting our exploratory data analysis. We also visualise some general features to show the relationship between the variables. Finally, we hypothesised some cases and tried to prove them visually.

## References 
1. "https://www.kaggle.com/code/fpolcari/spotify-songs-popularity-analysis"

2. "https://www.linkedin.com/pulse/spotify-telugu-playlist-analysis-using-r-amrutha-killada/"

3.  "https://github.com/kayamoon/spotify-music-data-analysis"

4.  "https://www.kaggle.com/code/trixietacung/spotify-2010-2019-insight-analysis-with-r/report"

